---
title: "Football match prediction using Correlated features and the bivariate Poisson Distribution."
author: "Suhishan Bhandari"
date: "December 25, 2025"
format:
  html: 
    code-fold: true
    code-tools: true
toc: TRUE
execute:
  message: false
  warning: false
---

```{r}
#| label: load packages
#| include: false
packages <- c("tidyverse","cmdstanr","bayesplot","loo","posterior","qs", "patchwork")
lapply(packages, library, character.only = TRUE)

```

Continuing from part 1, I want to write two more models, with the grand aim being to compare all four models in terms of their estimates as well as their predictability and out-of-sample deviance.

# Correlated Features Model
The third model that we are going to consider is a correlated features model, whereby a team's attack and defense strengths are correlated; the rationale is that teams who are good at attacking may also be good at defending (possible reasons being a.: attacking teams's higher possession may mean that the opposition team don't have enough of the ball to attack or b. financial discrepancies mean that teams higher up the table hire both better attackers and defenders and in general, the goal of football is to score goal on the one end and not concede on the other so the correlation between attack and defense is baked into the game)

Given this correlated rationale, the statistical model looks something like this:

$$
\begin{align}
s1 \sim poisson(\theta_1) \\ 
s2 \sim poisson(\theta_2) \\
\theta_1 = exp(home + att[ht] - def[at]) \\
\theta_2 = exp(att[at] - def[ht]) \\
\\
\text{Correlated features}
\\
\begin{bmatrix} att \\ def \end{bmatrix} = multinormal(
\begin{bmatrix} \overline{att} \\ \overline{def}\end{bmatrix}, 
\textbf{sigma_team})
\end{align}
$$

where, **sigma_team** is a covariance matrix as follows:

$$
\begin{align}
\textbf{sigma_team} = \begin{bmatrix} \sigma_{att} & 0 \\ 0 & \sigma_{def} \end{bmatrix} \begin{bmatrix} 1 & \rho \\ \rho & 1\end{bmatrix} 
\begin{bmatrix} \sigma_{att} & 0 \\ 0 & \sigma_{def} \end{bmatrix}
\end{align}
$$

This model allows us to capture the correlation between attack and defence parameters across teams, and allows us to answer the afore-mentioned queries of whether teams that are good in attack also good in defense (spoiler, they are!). While this was visible from the attack and defense estimates plots in the previous blog, this is a much more formal and methodical approach to the analysis.

Now, let's move onto the Stan code:

``` stan
{stan}
#| label: m_corr stan model
#| eval: false

data {
    int<lower=0> N;
    int<lower=0> Nt;
    int<lower=0> Np;
    array[N] int<lower=0> s1;
    array[N] int<lower=0> s2;
    array[N] int<lower=0> ht;
    array[N] int<lower=0> at;

    //predictions
    array[Np] int<lower=0> htp; // home team index for prediction
    array[Np] int<lower=0> atp; //away team index for prediction
}
parameters{
    real home;
    vector[Nt] att;
    vector[Nt] def;

    // for correlated features.
    corr_matrix[2] Rho;
    vector<lower=0>[2] sigma_team;
    real att_bar;
    real def_bar;
}
transformed parameters {
   vector[N] theta_1;
   vector[N] theta_2;

   theta_1 = exp(home + att[ht] - def[at]);
   theta_2 = exp(att[at] - def[ht]);
}
model{
    home ~ normal(0, 0.1);
    att_bar ~ normal(0, 0.1);
    def_bar ~ normal(0, 0.1);
    sigma_team ~ exponential(1);
    Rho ~ lkj_corr(2);

    {
    vector[2] MU;
    array[20] vector[2] YY;
    MU = [att_bar, def_bar]';
    for (j in 1:Nt) YY[j] = [att[j], def[j]]';
    YY ~ multi_normal(MU, quad_form_diag(Rho, sigma_team));
    }

    s1 ~ poisson(theta_1);
    s2 ~ poisson(theta_2);
}

generated quantities {
  vector[Np] theta1p; // home team predicted score
  vector[Np] theta2p; // away team predicted score.
  array[Np] int s1p; // score prediction for home using theta1p.
  array[Np] int s2p; // score prediction for away using theta2p.
  vector[N] log_lik;
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);

   
  for (i in 1:N){
    log_lik[i] = poisson_lpmf(s1[i] | theta_1[i]) + poisson_lpmf(s2[i] | theta_2[i]);
   }
   
}



```

Let us now look at the estimates, specifically the correlation parameter `Rho`.

```{r}
#| include: false
#| label : data cleaning and stuff

data <-  read.csv("data/premfull23_24.csv")

data = data %>% select(HomeTeam, HomeGoals, AwayGoals, AwayTeam) %>% 
  rename(
    Home = HomeTeam,
    s1 = HomeGoals,
    s2 = AwayGoals,
    Away = AwayTeam
  )

# Recode team names.
lookup <- c(
     "Brighton and Hove Albion" = "Brighton",
     "Manchester Utd" = "Manchester United",
     "Newcastle Utd" = "Newcastle United",
     "Nott'ham Forest" = "Nottingham Forest",
     "Sheffield Utd" = "Sheffield United",
     "Tottenham Hotspur" = "Tottenham",
     "West Ham United" = "West Ham",
     "Wolverhampton Wanderers" = "Wolves"
)

data = data %>% mutate(
  Home = recode(Home, !!!lookup),
  Away = recode(Away, !!!lookup)
)


teams = unique(data$Home) # Team names
ng = nrow(data) # number of games
nt = length(unique(data$Home)) # number of teams
# Home team index
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g]))) 
# Away team index
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g]))) 
np = 10 # Number of games to be predicted i.e. test sample length
ngd = ng - np # Number of games used to train i.e. train sample length.

```

```{r}
#| warning: false
#| label: cmdstan m_corr model
m_corr <- cmdstan_model("code/correlated_features.stan")
dat_corr <-  list(
  Nt = nt,
  N = ngd,
  Np = np,
  ht = ht[1:ngd],
  at = at[1:ngd],
  s1 = data$s1[1:ngd],
  s2 = data$s2[1:ngd],
  htp = ht[(ngd+1) : ng],
  atp = at[(ngd+1) : ng]
)

m_corr_fit <- m_corr$sample(data = dat_corr, chains = 4, parallel_chains = 4,
refresh = 500 )
 
```

Plot the attack and defense parameter picture:

```{r}
#| label: plot the correlated features.

m_corr_ad <- m_corr_fit$draws(variables = c("att", "def", "att_bar", "def_bar", "Rho"), format = "df") %>%
  data.frame()

d_corr <- tibble(
  att = apply(m_corr_ad[1:20], 2, mean),
  att_lc = apply(m_corr_ad[1:20], 2, function(x) quantile(x, probs = 0.03)),
  att_uc = apply(m_corr_ad[1:20], 2, function(x) quantile(x, probs = 0.97)),
  def = apply(m_corr_ad[21:40], 2, mean),
  def_lc = apply(m_corr_ad[21:40], 2, function(x) quantile(x, probs = 0.03)),
  def_uc = apply(m_corr_ad[21:40], 2, function(x) quantile(x, probs = 0.97)),
  teams = teams
)


plot_corr <- d_corr %>% ggplot(aes(x = att, y = def))+
  geom_point(shape = 1, size = 3)+
  geom_errorbar(aes(xmin = att_lc, xmax = att_uc), alpha = 0.1)+
  geom_errorbar(aes(ymin = def_lc, ymax = def_uc), alpha = 0.1)+
  geom_vline(xintercept = mean(m_corr_ad$att_bar),
             linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = mean(m_corr_ad$def_bar), linetype = 2, alpha = 0.5)+
  geom_text(data = d_corr[c(1, 2,3, 10, 13, 16, 20),], aes(label = teams),
            hjust = -0.15, vjust = 0.15)+
  labs(x = "Attack Strength", y = "Defence Strength",
       title = "PL 2023/24 Teams Att and Def Strengths",
       subtitle = "Correlated Features Pooling Model (94% credible intervals)")+
    coord_cartesian(xlim = c(-0.8, 0.8), y = c(-0.8, 0.8))+
  theme_classic()

plot_corr
```

```{r}
#| label: Plot the Rho.
mean_rho = round(mean(m_corr_ad$Rho.1.2.), 3)

tibble(
  `Posterior Rho` = m_corr_ad$Rho.1.2.,
  `Prior Rho` = rethinking::rlkjcorr(4000, K = 2, eta = 2)[, 1, 2]
) |> pivot_longer(everything()) |> 
  ggplot(aes(x = value))+
  geom_vline(xintercept = mean_rho, linetype = 2)+
  annotate("text", x = mean_rho, y = 1.5, label = as.character(mean_rho), hjust = -0.5)+
  annotate("text", x = mean_rho, y = 1.5, 
  label = "Mean Correlation", vjust = 2.5, hjust = 0.33)+
  geom_density(aes(color = name, fill = name), linewidth = 1.5, alpha = 0.2) +
  labs(title = "Comparison between Prior and Posterio Rho Estimates from the Correlated features model")+
  theme_bw()
```


# Bivariate Poisson Model.


``` stan
{stan}

data {
  int<lower=0> N;
  int<lower=0> Nt;
  int<lower=0> Np;
  array[N] int<lower=0> s1;
  array[N] int<lower=0> s2;
  array[N] int<lower=0> ht;
  array[N] int<lower=0> at;

  array[Np] int<lower=0> htp;
  array[Np] int<lower=0> atp;
}

parameters {
  real home;
  vector[Nt] att;
  vector[Nt] def;
  real<lower=0> theta_3;
}

transformed parameters {
  vector[N] theta_1;
  vector[N] theta_2;
  theta_1 = exp(home + att[ht] - def[at]);
  theta_2 = exp(att[at] - def[ht]);
}

model {
  // priors.
  home ~ normal(0, 0.1);
  att ~ normal(0, 0.1);
  def ~ normal(0, 0.1);
  theta_3 ~ normal(0, 0.5);

  s1 ~ poisson(theta_1 + theta_3);
  s2 ~ poisson(theta_2 + theta_3);
}

generated quantities {
  vector[Np] theta1p; 
  vector[Np] theta2p; 
  array[Np] int s1p; 
  array[Np] int s2p; 
  vector[N] log_lik;
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p + theta_3);
  s2p = poisson_rng(theta2p + theta_3);

  for (i in 1:N){
    log_lik[i] = poisson_lpmf(s1[i] | theta_1[i] + theta_3) + poisson_lpmf(s2[i] | theta_2[i] + theta_3);
   }

}


```


```{r}
m_bivar <- cmdstan_model("code/bivariate_poisson.stan")
m_bivar_fit <- m_bivar$sample(data = dat_corr, chains = 4, parallel_chains = 8, 
refresh = 500 )
```




## Generate Predictions from each model and compare:


```{r}
#| include: false
score_matrix <- function(post, game){
  n_sample <- length(post[[game]])
  home_p <- table(post[[game]])/n_sample # Posterior proportions of scores
  away_p <- table(post[[game + 10]])/n_sample
  
  # Matrix of results.
  m = round(home_p %o% away_p, 3)
  return (m)
}

# Win/Loss/Draw Percentages.

result_prob <- function(post, game) {
  m = score_matrix(post, game)
  hw = sum(m[lower.tri(m)])
  d = sum(diag(m))
  aw = sum(m[upper.tri(m)])
  
  return(c(hw = hw, d = d, aw = aw))
}

```

Predictions:


```{r}
m_corr_post <- m_corr_fit$draws(variables = c("s1p","s2p"), format = "df")
pred_corr <- sapply(1:10, function(x) result_prob(m_corr_post, x)) %>% 
  data.frame() %>% t() 
colnames(pred_corr) <- c("hw_c", "d_c", "aw_c")


m_bivar_post <- m_bivar_fit$draws(variables = c("s1p", "s2p"), format = "df")
pred_bivar <- sapply(1:10, function(x) result_prob(m_bivar_post, x)) %>% 
  data.frame() %>% t()
colnames(pred_bivar) <- c("hw_b", "d_b", "aw_b")

cbind(data %>% tail(n = 10), pred_corr, pred_bivar) %>% as_tibble()

# bivariate poisson seems extremely skeptical of things.
```