## This document outlines the various function calls for generating bivariate poisson log probability mass function 

### Function 1: Brute force 

$$
F_{(\theta_1, \theta_2, \theta_3)}(x, y) = 
exp(-(\theta_1 + \theta_2 + \theta_3)) 
\frac{\theta_1^x}{x!}  \frac{\theta_2^y}{y!} 
\sum^{min(x, y)}_{i = 0}i!{x \choose i}{y \choose i}
(\frac{\theta_3}{\theta_1 \theta_2})^i
$$

```{r}
bivar_poisson_lpmf <- function(x, y, lambda) {
  l_1  <- lambda[1] ;l_2 <- lambda[2]; l_3 <- lambda[3]
  mn <- min(x, y)
  con <- -l_1 - l_2 - l_3 + (x * log(l_1)) + (y * log(l_2)) - lfactorial(x) - lfactorial(y) 

  f <- numeric(length(mn) + 1)
  for (k in 0:mn) {
    f[k+1] <- choose(x, k)  * choose(y, k) * factorial(k) * (l_3/ (l_1 * l_2)) ^ k
  }

  f <- log(sum(f))
  a <- con + f

  return (a)
}

bivar_poisson_lpmf(1, 2, 1, 2, 3)
```


## Function 2: Brute force using underlying independent poisson distributions

$$
F_{(\theta_1, \theta_2, \theta_3)}(x, y) = 
\sum^{min(x, y)}_{i = 0} (e^{-\theta_3}\frac{\theta_3^i}{i!})
(e^{-\theta_1}\frac{\theta_1^{x-i}}{(x-i)!})
(e^{-\theta_2}\frac{\theta_2^{y-i}}{(y-i)!})
$$

```{r}
# A different way to write the same thing. (A cleaner and simpler way)

bivar_poisson_lpmf_2 <- function(x,y, lambda) {
  l_1  <- lambda[1] ;l_2 <- lambda[2]; l_3 <- lambda[3]
  mn  <- min(x, y)
  f <- numeric(length(mn) + 1)

  for (k in 0:mn) {
    f[k + 1] <- dpois(k, l_3) * dpois(x-k, l_1) * dpois(y-k, l_2)
  }
  a <- log(sum(f))
  return (a)
}

bivar_poisson_lpmf_2(1, 2, c(1, 2, 3))
```


## Function 3: Optimized way through iteration and numerically stable for Stan.

The third function relies on factoring out factorials.

$$
\begin{equation}
F_{(\theta_1, \theta_2, \theta_3)}(x, y) = 
\exp(-(\theta_1 + \theta_2 + \theta_3))
\sum^{min(x, y)}_{i = 0} \frac{\theta_3^i}{i!}
\frac{\theta_1^{x-i}}{(x-i)!}
\frac{\theta_2^{y-i}}{(y-i)!}
\end{equation}
$$

If we only take the what we need to sum,

$$
\begin{equation}
\begin{aligned}
S_i =  \frac{\theta_3^i}{i!}
\frac{\theta_1^{x-i}}{(x-i)!}
\frac{\theta_2^{y-i}}{(y-i)!}
\\
S_{i-1} = \frac{\theta_3^{i-1}}{{i-1}!}
\frac{\theta_1^{x-i + 1}}{(x-i + 1)!}
\frac{\theta_2^{y-i + 1}}{(y-i + 1)!} \\

\frac{S_i}{S_{i-1}} = \frac{\theta_3}{\theta_1 \theta_2}
\frac{(x-i + 1) (y - i + 1)}{i}\\
\end{aligned}
\end{equation}
$$

If we take the log of the third expression:

$$
log(S_i) = log(S_{i - 1}) + log(x -i + 1) + log (y -i + 1) -log(i) + log(\frac{\theta_3}{\theta_1 \theta_2})
$$

```{r}
bivar_poisson_lpmf_3 <- function(x, y, lambda) {
    l_1 <- log(lambda[1]); l_2 <- log(lambda[2]); l_3 <- log(lambda[3]);
    mn <- min(x, y)

    f <- numeric()
    # when l_3 is 0
    f[1] <- dpois(x, exp(l_1), log = TRUE) + dpois(y, exp(l_2), log = TRUE) -
        exp(l_3) # log(exp(-exp(l_3)))
    if (mn > 0) {
        cons <- -l_1 - l_2 + l_3
        for (i in 1:mn) {
            f[i+1] <- f[i] + log(x-i+1) + log(y-i+1) + cons - log(i)
        }
    }
    a <- log(sum(exp(f)))
    return(a)
}
```


Let's check the runtime for all three functions.

```{r}

```
