---
title: "Premier League Match Predictions using Bayesian Hierarchical Models"
author: "Suhishan Bhandari"
date: "December 25, 2025"
format:
  html: 
    code-fold: true
    code-tools: true
toc: TRUE
execute:
  message: false
  warning: false
---

```{r}
#| label: load packages
#| include: false
packages <- c("tidyverse","cmdstanr","bayesplot","loo","posterior","qs", "patchwork")
lapply(packages, library, character.only = TRUE)

```

# Introduction

The blog aims to estimate every Premier League team's attack and defence strengths using their match scores from $37$ gameweeks (from the 2023/24 Premier League season). Using these strength estimates, I then predict the results of the 38th gameweek.

# Model

$$
\begin{align}
s1 \sim poisson(\theta_1) \\
s2 \sim poisson(\theta_2) \\
\theta_1 = exp(home + att[ht]-def[at])\\
\theta_2 = exp(att[at] - def[ht])
\end{align}
$$

$s1$ and $s2$ are home and away scores respectively for a given match, which are assumed to be Poisson processes with their respective $\theta$. Scores are dependent on whether the team plays at home, and the attack and defence strengths of the home team ($ht$) and the away team($at$).

## Data

```{r}
#| include: false

#--- Load the data and make it usable ---#

data <-  read.csv("data/premfull23_24.csv")

data = data %>% select(HomeTeam, HomeGoals, AwayGoals, AwayTeam) %>% 
  rename(
    Home = HomeTeam,
    s1 = HomeGoals,
    s2 = AwayGoals,
    Away = AwayTeam
  )

# Recode team names.
lookup <- c(
     "Brighton and Hove Albion" = "Brighton",
     "Manchester Utd" = "Manchester United",
     "Newcastle Utd" = "Newcastle United",
     "Nott'ham Forest" = "Nottingham Forest",
     "Sheffield Utd" = "Sheffield United",
     "Tottenham Hotspur" = "Tottenham",
     "West Ham United" = "West Ham",
     "Wolverhampton Wanderers" = "Wolves"
)

data = data %>% mutate(
  Home = recode(Home, !!!lookup),
  Away = recode(Away, !!!lookup)
)

```

In accordance with the mathematical model above, the data contains information on home and away team scores. Later, for inferential purposes, team names are indexed from $i = 1...20$.

```{r}
head(data)
```

```{r}
teams = unique(data$Home) # Team names
ng = nrow(data) # number of games
nt = length(unique(data$Home)) # number of teams
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g]))) # Home team index
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g]))) # Away team index

np = 10 # Number of games to be predicted i.e. test sample length
ngd = ng - np # Number of games used to train i.e. train sample length.


dat <-  list(
  nt = nt,
  ng = ngd,
  ht = ht[1:ngd],
  at = at[1:ngd],
  s1 = data$s1[1:ngd],
  s2 = data$s2[1:ngd],
  
  # for predictions
  np = np,
  htp = ht[(ngd+1):ng],
  atp = at[(ngd+1):ng]
)

```

## Model 1

The first model we will use for our inference is a complete pooling model where every team gets its own attack and defence parameters (different intercept for each) but none of these parameters are related to each other in any way. The stan code for the statistical model is shown below[^1]:

[^1]: This model is influenced from the hierarchical model presented in stan's official YouTube channel. ([Link to the video.](https://www.youtube.com/watch?v=dNZQrcAjgXQ))

``` stan
{stan}
#| eval: false
data {
  int<lower=0> nt; // number of teams (20)
  int<lower=0> ng; // number of games (370)
  int<lower=0> np; // number of games to predict (final 10 games)
  array[ng] int<lower=1> ht; //home team index
  array[ng] int<lower=1> at; // away team index
  array[ng] int<lower=0> s1; // home team score
  array[ng] int<lower=0> s2; // away team score
  
  array[np] int<lower=0> htp; // home team index for prediction
  array[np] int<lower=0> atp; //away team index for prediction
}

parameters{
  real home;
  vector[nt] att;
  vector[nt] def;
}

transformed parameters{
  vector[ng] theta1; // poisson lambda for home team
  vector[ng] theta2; //  poisson lambda for away team
  
  theta1 = exp(home + att[ht] - def[at]);
  theta2 = exp(att[at] - def[ht]);
}

model{
  //priors
  att ~ normal(0, 1);
  def ~ normal(0, 1);
  home ~ normal(0, 1);
  
  // likelihood
  s1 ~ poisson(theta1);
  s2 ~ poisson(theta2);
}

generated quantities {
  vector[np] theta1p; // home team predicted score
  vector[np] theta2p; // away team predicted score.
  
  vector[ng] log_lik; // a vector of log-likelihood per observation
  
  array[np] int s1p; // score prediction for home using theta1p.
  array[np] int s2p; // score prediction for away using theta2p.
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);
  
  
  // TODO generate a log-likelihood for WAIC loo-PSIS.
  
  for (i in 1:ng) {
    log_lik[i] = poisson_lpmf(s1[i] | theta1[i]) + poisson_lpmf(s2[i] | theta2[i]);
  }
}
```

### Prior Predictive Simulation

Before we venture into the realms of estimation and prediction, let's see the prior distribution of our parameters. Home Scores ($s1$) and Away Scores ($s2$) depend upon some average rate $\theta_1$ and $\theta_2$ respectively, which again depend upon a team's attack and defence strengths throughout the season. If parameters `home`, `att` and `def` are assumed to be normally distributed, and with $theta$ positivity restriction i.e. exponentiating so that $theta$s remain positive, the home and away scores ranges from $0$ to almost $25$ goals and beyond. This means that with these priors, the model, *a priori* believes that PL teams can score upto 25 goals in a game! (which is absurd). This is almost certainly a very flat prior with the only information being embedded that scores must be positive. However, given that we have 370 matches as sample size to aid in our inference, these priors will do for now. In a separate blog (maybe part 2), I will try to tweak and regularize these priors to see what changes.

```{r}
n <- 1000
prior <- tibble(
  home = rnorm(n, 0, 1),
  att = rnorm(n, 0, 1),
  def = rnorm(n, 0, 1 )
) %>% mutate(
  theta_1 = exp(home + att - def),
  theta_2 = exp(att - def)
) %>% mutate(
  s1 = rpois(n, lambda = theta_1),
  s2 = rpois(n, lambda = theta_2)
)
```

```{r}
#| warning: false
s1_plot_prior <- prior %>% ggplot(aes(x = s1))+
  geom_histogram(fill = "skyblue", color = 'black', bins = 20) +
  xlim(0, 25)+
  ylim(0, 300)+
  labs(x = "Home Score")+
  theme_classic()

s2_plot_prior <- prior %>% ggplot(aes(x = s2))+
  geom_histogram(fill = "red", color = 'black', alpha = 0.5, bins = 20) +  
  xlim(0, 25)+
  ylim(0, 300)+
  labs(x = "Away Score")+
  theme_classic()

s1_plot_prior
s2_plot_prior
```

## Model 1 Estimates

I compile the stan model with `cmdstanr`, which peruses and estimates the posterior using HMC, and I extract the posterior samples for $att$ and $def$ parameters for 20 teams. The relative $att$ and $def$ estimates for the 20 PL teams in 2023/24 is as follows:

```{r}
#| warning: false
#| output: false
m1 <- cmdstan_model("code/normal_model.stan")

# Running this only once
m1fit <- m1$sample(
  data = dat, chains = 8, parallel_chains = 8, refresh = 500, output_dir = "fits/csv"
)
qsave(x = m1fit, file = 'fits/normal_model.qs')

# Load the fit from saved stan modelfit object
m1fit <- qread("fits/normal_model.qs")

```

Plot:

```{r}
# Model 1 attack and defence parameters posterior.
m1ad <- m1fit$draws(variables = c("att", "def"), format = "df") %>%
  data.frame()

d_m1 <- tibble(
  att1 = apply(m1ad[1:20], 2, mean),
  att1_lc = apply(m1ad[1:20], 2, function(x) quantile(x, probs = 0.03)),
  att1_uc = apply(m1ad[1:20], 2, function(x) quantile(x, probs = 0.97)),
  def1 = apply(m1ad[21:40], 2, mean),
  def1_lc = apply(m1ad[21:40], 2, function(x) quantile(x, probs = 0.03)),
  def1_uc = apply(m1ad[21:40], 2, function(x) quantile(x, probs = 0.97)),
  teams = teams
)
  
plot_normal <- d_m1 %>% ggplot(aes(x = att1, y = def1))+
  geom_point(shape = 1, size = 3)+
  geom_errorbar(aes(xmin = att1_lc, xmax = att1_uc), alpha = 0.1)+
  geom_errorbar(aes(ymin = def1_lc, ymax = def1_uc), alpha = 0.1)+
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5)+
  geom_text(data = d_m1[c(1, 2,3, 10, 13, 16, 20),], aes(label = teams),
            hjust = -0.15, vjust = 0.15)+
  labs(x = "Attack Strength", y = "Defence Strength",
       title = "PL 2023/24 Teams Att and Def Strengths",
       subtitle = "Complete Pooling Model (94% credible intervals)")+
    coord_cartesian(xlim = c(-0.8, 0.8), y = c(-0.8, 0.8))+
  theme_classic()

plot_normal


```

As shown by the figure, the model does a good job of providing relative attack and defence strength estimates, with power clubs like Arsenal, Manchester City, and Liverpool high on both metrics and relegated clubs Burnley, Luton Town and Sheffield United relatively weaker. It must be pointed out that while the point estimates show relative differences, the credible intervals are wide and as such, to see whether there's a significant difference between any two teams, we must use the posterior contrast. Nonetheless, the model provides a general picture of relative strengths.

Using these relative attack and defence strength parameters, we can then estimate likely scores and win/lose/draw probabilities for the final gameweek. For this purpose, I extract posterior samples for the home and away prediction for the final gameweek $s1p$ and $s2p$. Stan provides the posterior of the prediction for us because of the way the `generated quantities` block is coded. We could have also gotten $s1p$ and $s2p$ outside stan, by using predictions of $\theta_{1p}$ and $\theta_{2p}$.

Then, I write two functions: a) `score_matrix` which returns a matrix of probabilties of each home and away score, and b) `result_prob` which uses this score matrix to return win, loss and draw probabilities.

```{r}
# Score matrix function for a game
#--- Working with posterior draws ---#
m1post = m1fit$draws(variables = c("s1p", "s2p"), format = "df")

score_matrix <- function(post, game){
  n_sample <- length(post[[game]])
  home_p <- table(post[[game]])/n_sample # Posterior proportions of scores
  away_p <- table(post[[game + 10]])/n_sample
  
  # Matrix of results.
  m = round(home_p %o% away_p, 3)
  return (m)
}

# Win/Loss/Draw Percentages.

result_prob <- function(post, game) {
  m = score_matrix(post, game)
  hw = sum(m[lower.tri(m)])
  d = sum(diag(m))
  aw = sum(m[upper.tri(m)])
  
  return(c(hw = hw, d = d, aw = aw))
}

```

For example, for gameweek 38's match between Arsenal and Everton, the model estimates the goal matrix and result probabilities as follows:

```{r}
score_matrix(m1post, 1)
result_prob(m1post, 1)
```

The score matrix shows that the most likely scoreline is Arsenal 2 - 0 Everton (highest probability of joint occurrence) and that Arsenal is $75\%$ likely to win the match. If we estimate result probabilties for all 10 games we have:

```{r}

pred <- sapply(1:10, function(x) result_prob(m1post, x)) %>% 
  data.frame() %>% t() 

cbind(data %>% tail(n=10), pred) %>% as_tibble()
```

The model estimates results for 8 matches correct, with varying certainty. The model gets it wrong on the final 2 games: Brighton vs. Man United and Crystal Palace vs. Aston Villa. This sheds a light into one of the limitations of the model (among its many limitations): the model does not take into account recent matches momentum, which Crystal Palace were enjoying very much from Oliver Glasner's new appointment and a string of good results before the final gameweek.

Can a hierarchical model do better?

## Model 2: Hierarchical Model

In a hierarchical (or multilevel model), we assume that our attack and defence strength parameters come from  population attack and defence strength i.e. instead of assuming that each team's attack and defence strengths being modelled and estimated independent of other teams, the parameters are now draws from a population parameter. For e.g. $att[1..20]$ takes as its prior $att_bar$ as mean and $att_tau$ as standard deviation. The benefit of using such a hierarchical model is that it aids in minimizing over-fitting, whereby, if a team once scored 8 goals in match, the model is more sceptical than the complete-pooling model. Multilevel model shrinks extreme attack and defence strengths towards the mean. This is reasonable given the unpredictability of football results, whereby even in a game between a high-ranked team and a low-ranked team, the saying goes that 'anything can happen in one football game'. Since it is our goal to predict the final gameweek, multilevel model incorporates our own sense of unpredictability and scepticism rather well. Multi-level models also squeeze a lot more information out of the data.


The stan code:
``` stan
{stan}
#| eval: false
data {
  int<lower=0> nt; // number of teams (20)
  int<lower=0> ng; // number of games (370)
  int<lower=0> np; // number of games to predict (final 10 games)
  array[ng] int<lower=1> ht; //home team index
  array[ng] int<lower=1> at; // away team index
  array[ng] int<lower=0> s1; // home team score
  array[ng] int<lower=0> s2; // away team score
  
  array[np] int<lower=0> htp; // home team index for prediction
  array[np] int<lower=0> atp; //away team index for prediction
}

parameters{
  real home;
  vector[nt] att;
  vector[nt] def;
  real att_bar;
  real def_bar;
  real<lower=0> att_tau;
  real<lower=0> def_tau;
}

transformed parameters{
  vector[ng] theta1; // poisson lambda for home team
  vector[ng] theta2; //  poisson lambda for away team
  
  theta1 = exp(home + att[ht] - def[at]);
  theta2 = exp(att[at] - def[ht]);
}

model{
  
  // high-level hyper priors.
  att_bar ~ normal(0, 0.1);
  att_tau ~ normal(0, 1);
  def_bar ~ normal(0, 0.1);
  def_tau ~ normal(0, 01);
  
  //low-level priors
  att ~ normal(att_bar, att_tau);
  def ~ normal(def_bar, def_tau);
  home ~ normal(0, 1);
  
  // likelihood
  s1 ~ poisson(theta1);
  s2 ~ poisson(theta2);
}

generated quantities {
  vector[np] theta1p; // home team predicted score
  vector[np] theta2p; // away team predicted score.
  
  vector[ng] log_lik; // a vector of log-likelihood per observation
  
  array[np] int s1p; // score prediction for home using theta1p.
  array[np] int s2p; // score prediction for away using theta2p.
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);
  
  
  // TODO generate a log-likelihood for WAIC loo-PSIS.
  
  for (i in 1:ng) {
    log_lik[i] = poisson_lpmf(s1[i] | theta1[i]) + poisson_lpmf(s2[i] | theta2[i]);
  }
  
}
```
Fit the model:
```{r}
#| warning: false
#| output: false
m2 <- cmdstan_model("code/multilevel_model.stan")
m2fit <- m2$sample(data = dat, chains = 8, parallel_chains = 8,
                   refresh = 500, output_dir = "fits/csv")

qsave(x = m2fit, file = "fits/multilevel_model.qs")
m2fit <- qread("fits/multilevel_model.qs")

```

Plot the attack and defence strengths:

```{r}
m2ad <- m2fit$draws(variables = c("att", "def", "att_bar", "def_bar"),  format = "df") %>% data.frame()

d_m2 <- tibble(
  att2 = apply(m2ad[1:20], 2, mean),
  att2_lc = apply(m2ad[1:20], 2, function(x) quantile(x, probs = 0.03)),
  att2_uc = apply(m2ad[1:20], 2, function(x) quantile(x, probs = 0.97)),
  def2 = apply(m2ad[21:40], 2, mean),
  def2_lc = apply(m2ad[21:40], 2, function(x) quantile(x, probs = 0.03)),
  def2_uc = apply(m2ad[21:40], 2, function(x) quantile(x, probs = 0.97)),
  teams = teams
)
  
plot_multilevel <- d_m2 %>% ggplot(aes(x = att2, y = def2))+
  geom_point(shape = 1, size = 3)+
  geom_errorbar(aes(xmin = att2_lc, xmax = att2_uc), alpha = 0.2)+
  geom_errorbar(aes(ymin = def2_lc, ymax = def2_uc), alpha = 0.2)+
  geom_vline(xintercept = mean(m2ad$att_bar),
             linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = mean(m2ad$def_bar), linetype = 2, alpha = 0.5)+
  geom_text(data = d_m2[c(1, 2,3, 10, 13, 16, 20),], aes(label = teams),
            hjust = -0.15, vjust = 0.15)+
  annotate("text", -0.75, -0.05, label = expression(mu[def]))+
  annotate("text", 0.15, 0.75, label = expression(mu[att]))+
  
  labs(x = "Attack Strength", y = "Defence Strength",
       title = "PL 2023/24 Teams Att and Def Strengths",
       subtitle = "Hierarchical Model")+
  coord_cartesian(xlim = c(-0.8, 0.8), y = c(-0.8, 0.8))+
  theme_classic()


plot_multilevel


```
The figure plots the relative attack and defence strengths of teams in the Premier League using a hierarchical model. While we are going to see a comparison between complete pooling estimates and hierarchical model estimates, we can identity. one major difference in that the estimates are much more bunched together. Let's see what this model predicts in terms of result probabilities.

Predictions:

```{r}
m2post <- m2fit$draws(variables = c("s1p", "s2p"), format = "df")

pred2 <- sapply(1:10, function(x) result_prob(m2post, x)) %>% 
  data.frame() %>% t()
colnames(pred2) <- c("hw2", "d2", "aw2")
cbind(data %>% tail(n = 10), pred2) %>% as_tibble()

```
While the hit-rate for the hierarchical model is the same as the complete pooling model (8 out 10 correct), we see that hierarchical model predictions are much more conservative. For example, while the normal model gave $82\%$ chance that Manchester City win their last home game, the multilevel model predicts it at $77\%$.

## Comparison

The figure shows the most important difference between a complete-pooling and hierarchical model, that the hierarchical model is sceptical of relatively extreme estimates and tend to shrink towards the mean. As is shown by the figure, the red points are much closer to $\mu_att$ and $\mu_def$ than their black (complete pooling counterparts).

```{r}
tibble(
  att = c(d_m1$att1, d_m2$att2),
  def = c(d_m1$def1, d_m2$def2),
  Model = factor(rep(c("Normal","Multilevel"), each = 20)),
  team = factor(rep(seq(1, 20), 2))
) %>% ggplot(aes(x = att, y = def))+
  geom_point(aes(color = Model), size = 3)+
  scale_color_manual(
    values = c(
      "Normal" = "black",
      "Multilevel" = "red"
    )
  )+
  geom_line(aes(group = team), linewidth = 0.6)+
  geom_vline(xintercept = mean(m2ad$att_bar),
             linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = mean(m2ad$def_bar), linetype = 2, alpha = 0.5)+
  annotate("text", -0.5, -0.05, label = expression(mu[def]))+
  annotate("text", 0.15, 0.75, label = expression(mu[att]))+
  labs(title = "Comparison between complete pooling and multilevel model")+
  theme_classic()




```


Prediction Comparison:

While the hit-rates for correct predictions were the same, a point to be noted is that the hierarchical model had Crystal Palace on much higher likelihood of win $34.7\%$ than the normal model $32\%$. In the second part of this blog, we will look at and compare the out-of-sample deviance between the two models, especially if the sample size is small.

```{r}
cbind(data %>% tail(n = 10),
      pred,
      pred2) %>% as_tibble()
```
