---
title: "Premier League Match Predictions using Bayesian Hierarchical Models"
author: "Suhishan Bhandari"
date: "December 25, 2025"

toc: TRUE
---

```{r,include=FALSE, message=FALSE}
packages <- c("tidyverse","cmdstanr","bayesplot","loo","posterior")
lapply(packages, library, character.only = TRUE)

```

# Introduction

The blog aims to estimate every Premier League team's attack and defence strengths using their match scores from $37$ gameweeks (2023/24 season). Using these strength estimates, I then predict the results of the 38th gameweek.

# Model

$$
\begin{align}
s1 \sim poisson(\theta_1) \\
s2 \sim poisson(\theta_2) \\
\theta_1 = exp(home + att[ht]-def[at])\\
\theta_2 = exp(att[at] - def[ht])
\end{align}
$$

$s1$ and $s2$ are home and away scores respectively for a given match, which are assumed to be Poisson processes with their respective $\theta$. Scores are dependent on whether the team plays at home, and the attack and defence strengths of home team ($ht$) and away team($at$).

## Data

```{r, include=FALSE}
#--- Load the data and make it usable ---#

data <-  read.csv("data/premfull23_24.csv")

data = data %>% select(HomeTeam, HomeGoals, AwayGoals, AwayTeam) %>% 
  rename(
    Home = HomeTeam,
    s1 = HomeGoals,
    s2 = AwayGoals,
    Away = AwayTeam
  )

# Recode team names.
lookup <- c(
     "Brighton and Hove Albion" = "Brighton",
     "Manchester Utd" = "Manchester United",
     "Newcastle Utd" = "Newcastle United",
     "Nott'ham Forest" = "Nottingham Forest",
     "Sheffield Utd" = "Sheffield United",
     "Tottenham Hotspur" = "Tottenham",
     "West Ham United" = "West Ham",
     "Wolverhampton Wanderers" = "Wolves"
)

data = data %>% mutate(
  Home = recode(Home, !!!lookup),
  Away = recode(Away, !!!lookup)
)

```

```{r}
head(data)
```

## Model 1

The first model we will use for our inference is a complete pooling model where every team gets its own attack and defence parameters.

``` stan
{stan}
#| eval: false
data {
  int<lower=0> nt; // number of teams (20)
  int<lower=0> ng; // number of games (370)
  int<lower=0> np; // number of games to predict (final 10 games)
  array[ng] int<lower=1> ht; //home team index
  array[ng] int<lower=1> at; // away team index
  array[ng] int<lower=0> s1; // home team score
  array[ng] int<lower=0> s2; // away team score
  
  array[np] int<lower=0> htp; // home team index for prediction
  array[np] int<lower=0> atp; //away team index for prediction
}

parameters{
  real home;
  vector[nt] att;
  vector[nt] def;
}

transformed parameters{
  vector[ng] theta1; // poisson lambda for home team
  vector[ng] theta2; //  poisson lambda for away team
  
  theta1 = exp(home + att[ht] - def[at]);
  theta2 = exp(att[at] - def[ht]);
}

model{
  //priors
  att ~ normal(0, 1);
  def ~ normal(0, 1);
  home ~ normal(0, 1);
  
  // likelihood
  s1 ~ poisson(theta1);
  s2 ~ poisson(theta2);
}

generated quantities {
  vector[np] theta1p; // home team predicted score
  vector[np] theta2p; // away team predicted score.
  
  vector[ng] log_lik; // a vector of log-likelihood per observation
  
  array[np] int s1p; // score prediction for home using theta1p.
  array[np] int s2p; // score prediction for away using theta2p.
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);
  
  
  // TODO generate a log-likelihood for WAIC loo-PSIS.
  
  for (i in 1:ng) {
    log_lik[i] = poisson_lpmf(s1[i] | theta1[i]) + poisson_lpmf(s2[i] | theta2[i]);
  }
}
```



