---
title: "Premier League Match Predictions using Bayesian Hierarchical Models"
author: "Suhishan Bhandari"
date: "December 25, 2025"
toc: TRUE
---

```{r,include=FALSE, message=FALSE}
packages <- c("tidyverse","cmdstanr","bayesplot","loo","posterior")
lapply(packages, library, character.only = TRUE)

```

# Introduction

The blog aims to estimate every Premier League team's attack and defence strengths using their match scores from $37$ gameweeks (2023/24 season). Using these strength estimates, I then predict the results of the 38th gameweek.

# Model

$$
\begin{align}
s1 \sim poisson(\theta_1) \\
s2 \sim poisson(\theta_2) \\
\theta_1 = exp(home + att[ht]-def[at])\\
\theta_2 = exp(att[at] - def[ht])
\end{align}
$$

$s1$ and $s2$ are home and away scores respectively for a given match, which are assumed to be Poisson processes with their respective $\theta$. Scores are dependent on whether the team plays at home, and the attack and defence strengths of home team ($ht$) and away team($at$).

## Data

```{r, include=FALSE}
#--- Load the data and make it usable ---#

data <-  read.csv("data/premfull23_24.csv")

data = data %>% select(HomeTeam, HomeGoals, AwayGoals, AwayTeam) %>% 
  rename(
    Home = HomeTeam,
    s1 = HomeGoals,
    s2 = AwayGoals,
    Away = AwayTeam
  )

# Recode team names.
lookup <- c(
     "Brighton and Hove Albion" = "Brighton",
     "Manchester Utd" = "Manchester United",
     "Newcastle Utd" = "Newcastle United",
     "Nott'ham Forest" = "Nottingham Forest",
     "Sheffield Utd" = "Sheffield United",
     "Tottenham Hotspur" = "Tottenham",
     "West Ham United" = "West Ham",
     "Wolverhampton Wanderers" = "Wolves"
)

data = data %>% mutate(
  Home = recode(Home, !!!lookup),
  Away = recode(Away, !!!lookup)
)

```

```{r}
head(data)

```

```{r, include=FALSE}
teams = unique(data$Home) # Team names
ng = nrow(data) # number of games
nt = length(unique(data$Home)) # number of teams
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g]))) # Home team index
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g]))) # Away team index

np = 10 # Number of games to be predicted i.e. test sample length
ngd = ng - np # Number of games used to train i.e. train sample length.


dat <-  list(
  nt = nt,
  ng = ngd,
  ht = ht[1:ngd],
  at = at[1:ngd],
  s1 = data$s1[1:ngd],
  s2 = data$s2[1:ngd],
  
  # for predictions
  np = np,
  htp = ht[(ngd+1):ng],
  atp = at[(ngd+1):ng]
)

```

## Model 1

The first model we will use for our inference is a complete pooling model where every team gets its own attack and defence parameters (different intercept for each) but none of these parameters are related to each other in any way. The stan code for the statistical model is shown below:

``` stan
{stan}
#| eval: false
data {
  int<lower=0> nt; // number of teams (20)
  int<lower=0> ng; // number of games (370)
  int<lower=0> np; // number of games to predict (final 10 games)
  array[ng] int<lower=1> ht; //home team index
  array[ng] int<lower=1> at; // away team index
  array[ng] int<lower=0> s1; // home team score
  array[ng] int<lower=0> s2; // away team score
  
  array[np] int<lower=0> htp; // home team index for prediction
  array[np] int<lower=0> atp; //away team index for prediction
}

parameters{
  real home;
  vector[nt] att;
  vector[nt] def;
}

transformed parameters{
  vector[ng] theta1; // poisson lambda for home team
  vector[ng] theta2; //  poisson lambda for away team
  
  theta1 = exp(home + att[ht] - def[at]);
  theta2 = exp(att[at] - def[ht]);
}

model{
  //priors
  att ~ normal(0, 1);
  def ~ normal(0, 1);
  home ~ normal(0, 1);
  
  // likelihood
  s1 ~ poisson(theta1);
  s2 ~ poisson(theta2);
}

generated quantities {
  vector[np] theta1p; // home team predicted score
  vector[np] theta2p; // away team predicted score.
  
  vector[ng] log_lik; // a vector of log-likelihood per observation
  
  array[np] int s1p; // score prediction for home using theta1p.
  array[np] int s2p; // score prediction for away using theta2p.
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);
  
  
  // TODO generate a log-likelihood for WAIC loo-PSIS.
  
  for (i in 1:ng) {
    log_lik[i] = poisson_lpmf(s1[i] | theta1[i]) + poisson_lpmf(s2[i] | theta2[i]);
  }
}
```

## Model 1 Estimates

Extracting draws from the posterior:

```{r, message=FALSE, warning=FALSE}
m1 <- cmdstan_model("code/normal_model.stan")

m1fit <- m1$sample(
  data = dat, chains = 8, parallel_chains = 8, refresh = 500
)

#--- Working with posterior draws ---#
m1post = m1fit$draws(variables = c("s1p", "s2p"), format = "df")

```

Plot:

```{r}
# Model 1 attack and defence parameters posterior.
m1ad <- m1fit$draws(variables = c("att", "def"), format = "df") %>%
  data.frame()

d_m1 <- tibble(
  att1 = apply(m1ad[1:20], 2, mean),
  att1_lc = apply(m1ad[1:20], 2, function(x) quantile(x, probs = 0.03)),
  att1_uc = apply(m1ad[1:20], 2, function(x) quantile(x, probs = 0.97)),
  def1 = apply(m1ad[21:40], 2, mean),
  def1_lc = apply(m1ad[21:40], 2, function(x) quantile(x, probs = 0.03)),
  def1_uc = apply(m1ad[21:40], 2, function(x) quantile(x, probs = 0.97)),
  teams = teams
)
  
d_m1 %>% ggplot(aes(x = att1, y = def1))+
  geom_point(shape = 1, size = 3)+
  geom_errorbar(aes(xmin = att1_lc, xmax = att1_uc), alpha = 0.1)+
  geom_errorbar(aes(ymin = def1_lc, ymax = def1_uc), alpha = 0.1)+
  geom_vline(xintercept = 0, linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = 0, linetype = 2, alpha = 0.5)+
  geom_text(data = d_m1[c(1, 2,3, 10, 13, 16, 20),], aes(label = teams),
            hjust = -0.15, vjust = 0.15)+
  labs(x = "Attack Strength", y = "Defence Strength",
       title = "PL 2023/24 Teams Att and Def Strengths",
       subtitle = "Complete Pooling Model (94% credible intervals)")+
    coord_cartesian(xlim = c(-0.8, 0.8), y = c(-0.8, 0.8))+
  theme_classic()


```

These attack and defence estimates produce gameweek 38 predictions as follows:

```{r}
# Score matrix function for a game

score_matrix <- function(post, game){
  n_sample <- length(post[[game]])
  home_p <- table(post[[game]])/n_sample # Posterior proportions of scores
  away_p <- table(post[[game + 10]])/n_sample
  
  # Matrix of results.
  m = round(home_p %o% away_p, 4)
  return (m)
}

# Win/Loss/Draw Percentages.

result_prob <- function(post, game) {
  m = score_matrix(post, game)
  hw = sum(m[lower.tri(m)])
  d = sum(diag(m))
  aw = sum(m[upper.tri(m)])
  
  return(c(hw = hw, d = d, aw = aw))
}

pred <- sapply(1:10, function(x) result_prob(m1post, x)) %>% 
  data.frame() %>% t() 

cbind(data %>% tail(n=10), pred)
```

## Model 2: Hierarchical Model

``` stan
{stan}
#| eval: false
data {
  int<lower=0> nt; // number of teams (20)
  int<lower=0> ng; // number of games (370)
  int<lower=0> np; // number of games to predict (final 10 games)
  array[ng] int<lower=1> ht; //home team index
  array[ng] int<lower=1> at; // away team index
  array[ng] int<lower=0> s1; // home team score
  array[ng] int<lower=0> s2; // away team score
  
  array[np] int<lower=0> htp; // home team index for prediction
  array[np] int<lower=0> atp; //away team index for prediction
}

parameters{
  real home;
  vector[nt] att;
  vector[nt] def;
  real att_bar;
  real def_bar;
  real<lower=0> att_tau;
  real<lower=0> def_tau;
}

transformed parameters{
  vector[ng] theta1; // poisson lambda for home team
  vector[ng] theta2; //  poisson lambda for away team
  
  theta1 = exp(home + att[ht] - def[at]);
  theta2 = exp(att[at] - def[ht]);
}

model{
  
  // high-level hyper priors.
  att_bar ~ normal(0, 0.1);
  att_tau ~ normal(0, 1);
  def_bar ~ normal(0, 0.1);
  def_tau ~ normal(0, 01);
  
  //low-level priors
  att ~ normal(att_bar, att_tau);
  def ~ normal(def_bar, def_tau);
  home ~ normal(0, 1);
  
  // likelihood
  s1 ~ poisson(theta1);
  s2 ~ poisson(theta2);
}

generated quantities {
  vector[np] theta1p; // home team predicted score
  vector[np] theta2p; // away team predicted score.
  
  vector[ng] log_lik; // a vector of log-likelihood per observation
  
  array[np] int s1p; // score prediction for home using theta1p.
  array[np] int s2p; // score prediction for away using theta2p.
  
  theta1p = exp(home + att[htp] - def[atp]);
  theta2p = exp(att[atp] - def[htp]);
  
  s1p = poisson_rng(theta1p);
  s2p = poisson_rng(theta2p);
  
  
  // TODO generate a log-likelihood for WAIC loo-PSIS.
  
  for (i in 1:ng) {
    log_lik[i] = poisson_lpmf(s1[i] | theta1[i]) + poisson_lpmf(s2[i] | theta2[i]);
  }
  
}
```

```{r}
m2 <- cmdstan_model("code/multilevel_model.stan")
m2fit <- m2$sample(data = dat, chains = 8, parallel_chains = 8,
                   refresh = 500)

m2post <- m2fit$draws(variables = c("s1p", "s2p"), format = "df")

```

Plot the attack and defence strengths:

```{r}
m2ad <- m2fit$draws(variables = c("att", "def", "att_bar", "def_bar"),  format = "df") %>% data.frame()

d_m2 <- tibble(
  att2 = apply(m2ad[1:20], 2, mean),
  att2_lc = apply(m2ad[1:20], 2, function(x) quantile(x, probs = 0.03)),
  att2_uc = apply(m2ad[1:20], 2, function(x) quantile(x, probs = 0.97)),
  def2 = apply(m2ad[21:40], 2, mean),
  def2_lc = apply(m2ad[21:40], 2, function(x) quantile(x, probs = 0.03)),
  def2_uc = apply(m2ad[21:40], 2, function(x) quantile(x, probs = 0.97)),
  teams = teams
)
  
d_m2 %>% ggplot(aes(x = att2, y = def2))+
  geom_point(shape = 1, size = 3)+
  geom_errorbar(aes(xmin = att2_lc, xmax = att2_uc), alpha = 0.2)+
  geom_errorbar(aes(ymin = def2_lc, ymax = def2_uc), alpha = 0.2)+
  geom_vline(xintercept = mean(m2ad$att_bar),
             linetype = 2, alpha = 0.5)+
  geom_hline(yintercept = mean(m2ad$def_bar), linetype = 2, alpha = 0.5)+
  geom_text(data = d_m2[c(1, 2,3, 10, 13, 16, 20),], aes(label = teams),
            hjust = -0.15, vjust = 0.15)+
  annotate("text", -0.75, -0.05, label = expression(mu[def]))+
  annotate("text", 0.15, 0.75, label = expression(mu[att]))+
  
  labs(x = "Attack Strength", y = "Defence Strength",
       title = "PL 2023/24 Teams Att and Def Strengths",
       subtitle = "Hierarchical Model")+
  coord_cartesian(xlim = c(-0.8, 0.8), y = c(-0.8, 0.8))+
  theme_classic()



```

Predictions:

```{r}
pred2 <- sapply(1:10, function(x) result_prob(m2post, x)) %>% 
  data.frame() %>% t()

cbind(data %>% tail(n = 10), pred2)

```
